A Data-Driven Framework for Neural Field Modelling

### 2010-11-16 ltx/first_order_framework/LaTeX_Manuscript.tex
more edits from David to discussion

\subsection{Implications for Experimental Design}
As seen in Figures~\ref{fig:Figure7} and~\ref{fig:Figure10}, the results indicate accurate estimation for the connectivity kernel parameters, where the mean parameter estimates are in good agreement with the actual values and the true kernel lies within the 95~\% confidence intervals of the estimates. The estimates of the connectivity kernel parameters show relatively broad distributions, which can be expected considering the system's input is modelled as a random disturbance, and the observations are corrupted by noise. To overcome the effect of noise, the use of a controlled input stimulus should be considered when applying the estimation algorithm to electrophysiological recordings. Given knowledge of the input and the appropriate model, it is expected that the variance of the parameter estimates will decrease. Even when using an input, our results imply that we should only expect to learn distributions of parameters, rather than uncover a `true' parameter set. By designing an input signal carefully and using an evoked-potential paradigm, where stimuli are repeated many times, data could be epoched into (and averaged over) a sufficient number of trials (realisations), where each trial is a response to a stimulus, such that a distribution of the parameter estimates can be formed.

The frequency analysis specify the minimum sensor arrangement that is required to capture the dominant cortical dynamics. As pointed out in numerous publications by Freeman and colleagues (see \cite{Freeman1987} for example), a detailed description of the spatial properties of responses to stimuli is essential in understanding cortical dynamics. Importantly, the spatial frequency analysis tools presented in this paper can be used to design electrophysiological recording systems to avoid spatial aliasing, ensuring the essential spatial properties can be recovered. In addition, the frequency analysis highlights the relationship between the spatial width of the sensors and the information in the data. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density electrodes~\cite{Brinkmann2009}.

\subsection{Computational Complexity}
The closest competitor to our `grey-box' integro-difference equation based approach is perhaps the multivariate auto-regressive (MVAR) model, which is a `black-box' approach. Examples of MVAR models for estimating functional connectivity include Granger causality~\cite{Hesse2003}, the direct transfer function~\cite{Kaminski1991} and partial directed coherence~\cite{Sameshima1999}. A distinct advantage of the integro-difference equation based technique proposed in this current study is that the number of parameters required to model the system is considerably less than a typical MVAR model. Furthermore, by using the neural field equations as a parametric form of a model, the parameters have physical meaning and, therefore, can be clinically relevant. An advantage of the MVAR-based methods is that they allow for anisotropic connectivity, which is present in the longer range cortico-cortical fibres.

Previous work from Schiff and Sauer~\cite{schiff2008kalman} demonstrated for the first time the applicability of the unscented Kalman filter (UKF) for state estimation of discrete (in space and time) nonlinear spatiotemporal systems. Within their framework, each discrete spatial location forms an element of the state vector to be estimated with the UKF. A limitation of this framework is that an increase in spatial resolution and/or the physical size of the neural field will result in an increased state dimension.

This, in turn, will lead to increases in uncertainty of the estimates. In addition, directly linking the complexity of the underlying system to the spatial resolution of the observation process implies that the system's dynamics become increasingly more complex the closer we look, which may not be an accurate modelling assumption.

The current study builds on Schiff and Sauer's approach~\cite{schiff2008kalman} by estimating a continuous field whose complexity is independent of the resolution of the observation process and spatial resolution of the model. Instead, the dimensionality is governed by the expected complexity of the underlying field, which we measure using the frequency analysis presented in Methods section. These methods provide a rigorous framework for determining the dimensionality of the system using inherent spatial properties of the field. This enables estimation of a larger field with a lower dimensional state vector than would be otherwise possible.

That the connectivity kernel is isotropic means the mixing effect of the kernel on the firing rates can be written as a convolution of the kernel with the membrane voltage field basis functions. This transformation greatly simplifies the computational complexity of the estimator, which is important for the estimator to run in a real-time scenario.

@@

% Do NOT remove this, even if you are not including acknowledgments
\section{Acknowledgments}
Dean Freestone would like to thank Dr. Mark van Rossum for hosting him while visiting the University of Edinburgh where this work was undertaken and Dr. Matthias Hennig for useful discussions and feedback. Parham Aram would also like to thank Andrew Hills for useful discussions and assistance with code and the generation of figures.

>>>>

\subsection{Implications for Experimental Design}
As seen in Figures~\ref{fig:Figure7} and~\ref{fig:Figure10}, the results indicate accurate estimation for the connectivity kernel parameters, where the mean parameter estimates are in good agreement with the actual values where the true kernels lie within the 95~\% confidence intervals of the kernel estimates. The estimates of the connectivity kernel parameters show relatively broad distributions, which can be expected considering the system's input is modelled as a random disturbance and the observations are corrupted by noise. To overcome the effect of noise, the use of a controlled input stimulus should be considered when applying the estimation algorithm to electrophysiological recordings. Given knowledge of the input and the appropriate model, it is expected that the variance of the parameter estimates would decrease. Even when using a known input, our results imply that we should only expect to learn distributions of parameters, rather than uncover a `true' parameter set. By designing an input signal carefully and using an evoked-potential paradigm where stimuli are repeated many times, recordings could be averaged over a sufficient number of trials (realisations), where each trial is a response to a stimulus, such that a distribution of the parameter estimates can be formed.

The frequency analysis specifies the minimum sensor arrangement that is required to capture the dominant cortical dynamics. As pointed out in numerous publications by Freeman and colleagues (see \cite{Freeman1987} for example), a detailed description of the spatial properties of responses to stimuli is essential in understanding cortical dynamics. Importantly, the spatial frequency analysis tools presented in this paper can be used to design electrophysiological recording systems to avoid spatial aliasing, ensuring the essential spatial properties can be recovered. In addition, the frequency analysis highlights the relationship between the spatial width of the sensors and the information in the data. This analysis is becoming increasingly relevant as recording systems are becoming more sophisticated with higher density electrodes~\cite{Brinkmann2009}.

\subsection{Computational Complexity}
The closest competitor to our `grey-box' integro-difference equation approach is perhaps the multivariate auto-regressive (MVAR) model, which is a `black-box' approach. Examples of MVAR models for estimating functional connectivity include Granger causality~\cite{Hesse2003}, the direct transfer function~\cite{Kaminski1991}, and partial directed coherence~\cite{Sameshima1999}. A distinct advantage of the integro-difference equation based technique proposed in this current study is that the number of parameters required to model the system is considerably less than a typical MVAR model. Furthermore, by using the neural field equations as a parametric form of a model, the estimated quantities have physical meaning and, therefore, can be clinically relevant. An advantage of the MVAR-based methods is that they allow for anisotropic connectivity, which is present in the longer range cortico-cortical fibres.

Previous work by Schiff and Sauer~\cite{schiff2008kalman} demonstrated for the first time the applicability of the unscented Kalman filter (UKF) for state estimation of discrete (in space and time) nonlinear spatiotemporal systems. Within their framework, each discrete spatial location formed an element of the state vector to be estimated with the UKF. A limitation of this framework was that an increase in spatial resolution and/or the physical size of the neural field resulted in an increased state dimension. This, in turn, leads to increases in uncertainty in the state estimates. In addition, directly linking the complexity of the underlying system to the spatial resolution of the observation process implies that the system's dynamics become increasingly more complex the closer we look, which may not be an accurate modelling assumption.

The current study builds on Schiff and Sauer's approach~\cite{schiff2008kalman} by estimating a continuous field whose complexity is independent of the resolution of the observation process and spatial resolution of the model. Instead, the dimensionality is governed by the expected complexity of the underlying field, which we measure using the frequency analysis presented in the Methods section. These methods provide a rigorous framework for determining the dimensionality of the system using inherent spatial properties of the field. This enables estimation of a larger field with a lower dimensional state vector than would be otherwise possible.

The isotropy of the connectivity kernel means that the mixing effect of the kernel on the firing rates can be written as a convolution of the kernel with the membrane voltage field basis functions. This transformation greatly simplifies the computational complexity of the estimator, which is important for the estimator to run in a real-time scenario.

@@

\section{Acknowledgments}
Dean Freestone would like to thank Dr. Mark van Rossum for hosting him while visiting the University of Edinburgh where this work was undertaken and Dr. Matthias Hennig for useful discussions and feedback. Parham Aram would also like to thank Andrew Hills for useful discussions and assistance with code and the generation of figures.
